---
title: "Customer Churn Analysis & Prediction Report"
author: "Abeer Gaber"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = TRUE, fig.align = 'center')
library(tidyverse)
library(caret)
library(ggplot2)
library(pROC)
```

## 1. Executive Summary
This report aims to investigate and identify the primary drivers of Customer Churn for a telecommunications provider. Using statistical modeling (Logistic Regression), we predict customers at high risk of leaving to enable targeted, proactive retention campaigns.

## 2. Data Loading & Pre-processing
In this step, customer records are loaded, cleaned, and categorical variables are encoded for statistical modeling:

```{r preprocessing}
data <- read.csv("D:/Data Analysis/customers_churn/Telco-Customer-Churn.csv")

data_cleaned <- data %>%
  mutate(
    Churn_Numeric = ifelse(Churn == "Yes", 1, 0),
    Churn = as.factor(Churn),
    Contract = as.factor(Contract),
    InternetService = as.factor(InternetService),
    TechSupport = as.factor(TechSupport)
  ) %>%
  filter(!is.na(MonthlyCharges))

glimpse(data_cleaned)
```

## 3. Visual Exploratory Data Analysis (EDA)

### A) Churn Rate Distribution by Contract Type
```{r plot_contract}
ggplot(data_cleaned, aes(x = Contract, fill = Churn)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(
    title = "Churn Rate Distribution by Contract Type",
    x = "Contract Type",
    y = "Percentage",
    fill = "Churn Status"
  ) +
  theme_minimal()
```

**Key Insight:** Over 40% of Month-to-Month contract holders churned, compared to less than 3% churn rate among 2-Year contract holders.

### B) Monthly Charges Distribution by Churn Status
```{r plot_charges}
ggplot(data_cleaned, aes(x = Contract, fill = Churn)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(
    title = "Churn Rate Distribution by Contract Type",
    x = "Contract Type",
    y = "Percentage",
    fill = "Churn Status"
  ) +
  theme_minimal()
```

**Key Insight:** Churned customers have a higher median monthly charge (~$74.40) compared to retained customers (~$61.20).

## 4. Predictive Modeling (Logistic Regression)
The dataset was split (80% Train / 20% Test) to train a Logistic Regression model to uncover key statistical predictors:

```{r modeling}
set.seed(123)
train_index <- createDataPartition(data_cleaned$Churn, p = 0.8, list = FALSE)
train_data  <- data_cleaned[train_index, ]
test_data   <- data_cleaned[-train_index, ]

churn_model <- glm(
  Churn_Numeric ~ tenure + MonthlyCharges + Contract + InternetService + TechSupport,
  data = train_data,
  family = binomial(link = "logit")
)

summary(churn_model)
```

### A) Feature Importance (New Addition)
```{r feature_importance}
importance <- as.data.frame(summary(churn_model)$coefficients)
importance$Feature <- rownames(importance)
importance <- importance[-1, ] 

ggplot(importance, aes(x = reorder(Feature, abs(`z value`)), y = `z value`, fill = `z value` > 0)) +
  geom_bar(stat = "identity", width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c"), labels = c("Reduces Churn", "Increases Churn")) +
  labs(
    title = "Key Predictors of Customer Churn",
    subtitle = "Based on Statistical Significance (z-value)",
    x = "Features / Drivers",
    y = "Impact Score (z-value)",
    fill = "Effect Type"
  ) +
  theme_minimal()
```

**Key Insight & Business Value:** Features with high positive z-values (like Monthly Charges and Fiber Optic Internet) significantly increase the likelihood of churn. Features with negative z-values (like long tenure and Tech Support) act as retention anchors, keeping customers loyal.

## 5. Model Evaluation (New Addition)
To ensure our predictions are reliable before taking business actions, we evaluate the model on the unseen Test data:

```{r evaluation}
test_preds <- predict(churn_model, newdata = test_data, type = "response")
test_pred_class <- as.factor(ifelse(test_preds >= 0.3, "Yes", "No")) 

conf_matrix <- confusionMatrix(test_pred_class, test_data$Churn, positive = "Yes")
print(conf_matrix$table)

roc_curve <- roc(test_data$Churn, test_preds)
cat("Model Accuracy:", round(conf_matrix$overall['Accuracy'] * 100, 2), "%\n")
cat("Model Sensitivity (Catching Churners):", round(conf_matrix$byClass['Sensitivity'] * 100, 2), "%\n")
cat("Area Under Curve (AUC):", round(auc(roc_curve), 2), "\n")

```

**Key Insight & Business Value:** Evaluating metrics ensures the business doesn't waste budget on wrong predictions. A higher sensitivity means we correctly capture the majority of customers who are truly planning to leave.

## 6. Risk Predictions & Export
We generate individual churn probabilities for all customers and group them into actionable risk tiers:

```{r predictions}
data_cleaned$Churn_Probability <- predict(churn_model, newdata = data_cleaned, type = "response")

data_cleaned <- data_cleaned %>%
  mutate(
    Risk_Level = case_when(
      Churn_Probability >= 0.7 ~ "High Risk",
      Churn_Probability >= 0.3 ~ "Medium Risk",
      TRUE ~ "Low Risk"
    )
  )

write.csv(data_cleaned, "D:/Data Analysis/customer_churn/customers_with_predictions.csv", row.names = FALSE)

```

### Customer Distribution by Predictive Risk Level:
```{r plot_risk}
ggplot(data_cleaned, aes(x = Risk_Level, fill = Risk_Level)) +
  geom_bar(width = 0.5) +
  scale_fill_manual(values = c("#e74c3c", "#2ecc71", "#f1c40f")) +
  labs(
    title = "Customer Distribution by Predictive Risk Level",
    x = "Risk Level",
    y = "Customer Count"
  ) +
  theme_minimal()

```

## Strategic Business Recommendations
* **Contract Migration Strategy:** Offer a 10% discount incentive to convert Month-to-Month users into 1-Year or 

2-Year contracts.
* **Onboarding Support Program:** Focus customer support and onboarding resources on new customers during their 

first 6 months.
* **Targeted Retention Campaigns:** Export the High Risk list directly from `customers_with_predictions.csv` to launch targeted retention offers.
